Precision calculations for cosmic shear

### 2017-02-08 limber2.tex
MK edit text
Some updates to the intro

The measurement of weak gravitational lensing by large-scale structures
provides a powerful cosmological probe of dark matter, dark energy, and
modifications to gravity.  As such it is the primary science goal of several
current (KiDS, HSC, DES) and future (Euclid, LSST, WFIRST) large-scale surveys.
Interest in the results from these surveys is high as statistically significant
deviations have been found between the cosmological parameter constraints from
the CMB Planck experiment \citep{2015arXiv150201589P} in comparison to weak
lensing constraints from both the Kilo-Degree Survey \citep[KiDS;][]{KiDS-450}
and the Canada-France-Hawaii Telescope Lensing Survey
\citep[CFHTLenS;][]{joudaki/etal:2016}.  If the source of this tension is not
a result of so-far unconsidered sources of systematic errors in one or all
experiments, extensions to the standard flat $\Lambda$CDM cosmological models
need to be considered. \citet{joudaki/etal:2017} have shown, for example, that
the tension can be resolved with an evolving dark energy model.

In the era of the upcoming large surveys that will provide measurements of
cosmic shear with unprecedented precision, one needs to revisit the theoretical
model predictions of the observables to ensure that the accuracy of the
models meet the upcoming data. In this paper, we examine the widely used Limber
approximation for the projected weak-lensing power spectrum. We consider
spherical coordinates and the flat-sky approximation, and compute the full
projection as well as first- and second-order Limber approximation. We show
that the latter is an excellent approximation of the full projection. Since it
involves only 1D integrals over the matter power spectrum, it is very fast to
calculate numerically and can readily employed in Monte-Carlo sampling methods
to obtain precision constraints on cosmological parameters.

This paper is organised as follows.  In section~\ref{sec:wl} we provide a pedagogical introduction to weak gravitational lensing theory, projections and power spectra for the flat-sky and spherical cases.  In section~\ref{sec:L2} we derive weak lensing observables using a second-order limber approximation first introduced by \citet{2008PhRvD..78l3506L}.  We compare the shear power spectrum and commonly-used two-point shear correlation function for a range of different approximations in section~\ref{sec:comp}, providing cosmological constraints for each case using CFHTLenS data from \citet{CFHTLenS-2pt-notomo} in section~\ref{sec:cfhtlens}.  This paper draws from several sources of literature which have previously reviewed the accuracy of the Limber approximation for weak lensing studies, namely \citet{2008PhRvD..78d3002S,2012PhRvD..86b3001B, 2012MNRAS.422.2854G}.

@@

With this approximation, the $k$-independent term factors out
from (\ref{eq:f_LA08}), and we define the separated kernel function $f_{\rm s}$ as
%
\begin{equation}
  f_{\rm s}(\chi) = \sqrt{P_{\rm m}(k)} D_+(\chi) a^{-1}(\chi) \chi^{-3/2} q(\chi).
  \label{eq:f_LA08_s}
\end{equation}
%
and does not need to be considered when taking the derivatives with respect to comoving distances.
The second-order Limber term of the shear power spectrum can then be simplified to
%
\begin{align}
  C^\gamma_{{\rm L2}, ij}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ellbar^2(\ell, 2)}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} a^{-2} P_{\rm m}\left(\frac\nu\chi; \chi\right)
    %\nonumber \\
    %& \times
    \frac {a(\chi)}{2 D_+(\chi)} \chi^{3/2}
    \left[ q^{}_i f^{\prime\prime}_{{\rm s} j} + f^{\prime\prime}_{{\rm s}i} q^{}_j
      + \frac{\chi}{3} \left( q^{}_i f^{\prime\prime\prime}_{{\rm s}j} + f^{\prime\prime\prime}_{{\rm s}i} q^{}_j
      \right)
    \right](\chi)
  \label{eq:C_ell_limber2_dr_s}
\end{align}
%
We compute the numerical higher derivatives as follows. The functions $f_{\rm
s}(\chi)$ are fitted as power laws with index $\approx 1.5$. The fit is
performed between $\chi_{\rm min} = 0.001 {\rm Mpc}/h$ and $\chi_{\rm max} =
500 \mbox{Mpc}/h$. At larger comoving distances the kernel decreases faster
than a power law, so we exclude this range from the fit. Even though the
derivaties are larger due to the steeper decline, the associated errors are
very small since those scales are down-weighed by the kernel function $f_{\rm
s}$ itself. At $\chi = 500 \mbox{Mpc}/h$ the filter function is less than
$10^{-4}$ of its value of unity at Mpc$/h$.

>>>>

The measurement of weak gravitational lensing by large-scale structures
provides a powerful cosmological probe of dark matter, dark energy, and
modifications to gravity.  As such it is the primary science goal of several
current (KiDS, DES, HSC\footnote{KiDS: \href={http://kids.strw.leidenuniv.nl}, DES: \href={http://www.darkenergysurvey.org}, HSC: \href={http://hsc.mtk.nao.ac.jp/ssp/}})
and future (Euclid, LSST, WFIRST\footnote{Euclid: \href={http://sci.esa.int/euclid}, LSST: \href={http://www.lsst.org}, WFIRST: \href={http://wfirst.gsfc.nasa.gov}}) surveys.
%(Kilo-Degree Survey, KiDS \citet{kuijken/etal:2015}, Dark Energy Survey, DES, \citet{abbott/etal:2016}, Hyper-Suprime Camera Survey, HSC\footnote{HSC; \href={http://hsc.mtk.nao.ac.jp/ssp/}}) and future (Euclid, \citet{2011arXiv1110.3193L}, LSST, \citet{chang/etal:2013}, WFIRST\footnote{WFIRST; \href={http://wfirst.gsfc.nasa.gov}}) large-scale surveys.
Interest in the results from these surveys is high as statistically significant
deviations have been found between the cosmological parameter constraints from
the CMB Planck experiment \citep{2015arXiv150201589P} in comparison to weak
lensing constraints from both the Kilo-Degree Survey \citep[KiDS;][]{KiDS-450}
and the Canada-France-Hawaii Telescope Lensing Survey
\citep[CFHTLenS;][]{joudaki/etal:2016}.  If the source of this tension is not
a result of so-far unconsidered sources of systematic errors in one or all
experiments, extensions to the standard flat $\Lambda$CDM cosmological models
need to be considered. \citet{joudaki/etal:2017} have shown, for example, that
the tension can be resolved with an evolving dark energy model.

In the era of the upcoming large-scale surveys that will provide measurements of
cosmic shear with unprecedented precision, one needs to revisit the theoretical
predictions of the observables to ensure that the accuracy of the
models meet the accuracy of the upcoming data. In this paper, we examine the widely used Limber
approximation for the projected weak-lensing power spectrum. We consider
spherical coordinates and the flat-sky approximation, and compute the full
projection as well as first- and second-order Limber approximation. We show
that the second-order Limber approximation is an accurate representation of the full projection, with better than percent level precision for scale $\ell > 3$. 
Since this approximation involves only 1D integrals over the matter power spectrum, it is very fast to
calculate numerically and can readily employed in Monte-Carlo sampling methods
to obtain precision constraints on cosmological parameters.

This paper is organised as follows.  In section~\ref{sec:wl} we provide a pedagogical introduction to weak gravitational lensing theory, projections and power spectra for the flat-sky and spherical cases.  In section~\ref{sec:L2} we derive weak lensing observables using a second-order limber approximation first introduced by \citet{2008PhRvD..78l3506L}.  We compare the shear power spectrum and commonly-used two-point shear correlation function for the full solution to a range of different approximations in section~\ref{sec:comp}, providing cosmological constraints for each case using CFHTLenS data from \citet{CFHTLenS-2pt-notomo} in section~\ref{sec:cfhtlens}.  This paper draws from several sources of literature which have previously reviewed the accuracy of the Limber approximation for weak lensing studies, namely \citet{2008PhRvD..78d3002S,2012PhRvD..86b3001B, 2012MNRAS.422.2854G}.

@@

With this approximation, the $k$-independent term factors out from
(\ref{eq:f_LA08}), and we define the separated kernel function $f_{\rm s}$ as
%
\begin{equation}
  f_{\rm s}(\chi) = D_+(\chi) a^{-1}(\chi) \chi^{-3/2} q(\chi).
  \label{eq:f_LA08_s}
\end{equation}
%
and does not need to be considered when taking the derivatives with respect to comoving distances.
The second-order Limber term of the shear power spectrum can then be simplified to
%
\begin{align}
  C^\gamma_{{\rm L2}, ij}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ellbar^2(\ell, 2)}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} a^{-2} P_{\rm m}\left(\frac\nu\chi; \chi\right)
    %\nonumber \\
    %& \times
    \frac {a(\chi)}{2 D_+(\chi)} \chi^{3/2}
    \left[ q^{}_i f^{\prime\prime}_{{\rm s} j} + f^{\prime\prime}_{{\rm s}i} q^{}_j
      + \frac{\chi}{3} \left( q^{}_i f^{\prime\prime\prime}_{{\rm s}j} + f^{\prime\prime\prime}_{{\rm s}i} q^{}_j
      \right)
    \right](\chi)
  \label{eq:C_ell_limber2_dr_s}
\end{align}
%
We compute the numerical higher derivatives as follows. The functions $f_{\rm
s}(\chi)$ are fitted as power laws with index $\approx 1.5$. The fit is
performed between $\chi_{\rm min} = 0.001 {\rm Mpc}/h$ and $\chi_{\rm max} =
500 \mbox{Mpc}/h$. At larger comoving distances the kernel decreases faster
than a power law, so we exclude this range from the fit. Even though the
derivaties are larger due to the steeper decline, the associated errors are
very small since those scales are down-weighed by the kernel function $f_{\rm
s}$ itself. At $\chi = 500 \mbox{Mpc}/h$ the filter function is less than
$10^{-4}$ of its value of unity at Mpc$/h$.
